---
title: "*ezmmek*: Easy Michaelis-Menten Enzyme Kinetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ezmmek_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction

Enzyme assays are popular methods for measuring enzymatic activity. Such methods often involve the use of fluorescent and chromogenic substrate proxies. Based on traditional Michaelis-Menten kinetics, *ezmmek* contains a suite of functions that calibrate, load, and plot the activities of extracellular enzymes. *ezmmek* outputs plots and statistics relevant to standard curves and saturation curves. It also enables the user to customize plots to their particular datasets.

*ezmmek* provides two methods to calculate enzymatic activity. The first, as described by German et al. (2011) corrects for quenching  by considering the individual contributions of the different components (Method 1):
\
$$\tiny (1)~Activity~(nmol~g⁻¹~h⁻¹)~=~\frac{Net~Fluorescence~*~Buffer~Volume~(mL)}{Emission~Coefficient~*~ Homogenate~Volume~(mL)~*~Time~(h)~*~Soil~Mass~(g)}$$
\
$$\tiny (2)~Net~Fluorescence~=~\frac{Assay~-~Homogenate~Control}{Quench~Coefficient}~-~Substrate~Control~(mL)$$
\
$$\tiny (3)~Emission~Coeff.~(fluorescence~nmol⁻¹)~=~\frac{Standard~Curve~Slope~(in~presence~of~homogenate)~(\frac{Fluorescence}{\frac{nmol}{ml}})}{Assay~Volume~(mL)}$$
\
$$\tiny (4)~Quench~Coefficient~=~\frac{Slope~of~Standard~Curve~(in~presence~of~homogenate)}{Slope~of~Standard~Curve~(in~presence~of~buffer)}$$
\
\
The second method corrects for quenching by considering the bulk contribution of the different components (Method 2):
\
$$\tiny (5)~Activity~(nmol~g⁻¹~h⁻¹)~=~\frac{Concentration~of~Fluorophore~(nmol)}{Time~(h)~*~Soil~Mass~(g)}$$
\
$$\tiny (6)~Concentration~of~Fluorophore~(nmol)~=~\frac{Assay~Fluorescence~-~Intercept~of~Standard~Curve~(in~presence~of~homogenate)}{Slope~of~Standard~Curve~(in~presence~of~homogenate)}$$
\
\

## 2 Installation

The lastest versions of *ezmmek* are available through CRAN and GitHub.

### 2.1 CRAN

```{r, eval = FALSE}
install.packages("ezmmek")
```

### 2.2 GitHub


#### 2.2.1 Install and load *devtools* if needed
```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
```

#### 2.2.2 Install *ezmmek*
```{r, eval = FALSE}
install_github("ccook/ezmmek")
```

## 3 How to use

```{r, warning = FALSE}
library(ezmmek)
```

### 3.1 Check example datasets

#### 3.1.1 Standard curve data, Method 1
Contains time, concentration of substrate (sub.conc), replicate number (replicate), spectral data in presence of homogenate (spec.homo), and spectral data in presence of buffer (spec.buffer)

```{r, eval = FALSE}
d_std_g
```
\

#### 3.1.2 Saturation curve data, Method 1
Contains time, concentration of substrate (sub.conc), replicate number (replicate), and spectral data (spec), assay volume (assay.vol), substrate control (sub.control), buffer volume (buffer.vol), homogenate volume (homo.vol) and soil mass (soil.mass) 

```{r, eval = FALSE}
d_sat_g
```
\

#### 3.1.3 Standard curve data, Method 2 
Contains concentration of standard (std.conc) and spectral data (spec). 

```{r, eval = FALSE}
d_std
```
\

#### 3.1.4 Saturation curve data, Method 2 

Contains time, concentration of substrate (sub.conc), replicate number (replicate), and spectral data (spec).
    
```{r, eval = FALSE}
d_sat
```
\


### 3.2 Import datasets
Imported datasets must contain the same column headers as the example datasets provided here.

```{r, eval = FALSE}
your_data <- read.csv(file = "/your_path/your_data")
```

### 3.3 Functions

#### 3.3.1 Plot standard curve , Method 1
*g_std_curve()* plots spectral data vs standard concentration (in presence of both buffer and homogenate).  It reports linear model summary statistics. It creates a list output containing the summary statistics and standard curve plots.

```{r, fig.height = 4, fig.width = 5}
g_std_curve(d_std_g)
```
\

#### 3.3.2 Plot saturation curve, Method 1
*g_sat_curve()* converts the raw spectral data is used to calculate activity. The new dataframe contains the quench coefficients ('quench.coef), emission coefficients ('emission.coef'), homogenate corrected spectral values ('spec.homogenate.corrected'), and activities. 'g_sat_curve' plots the new dataframe with substrate concentration on the x-axis, and rate of reaction on the y-axis. It predicts and reports Vmax and Km values. It creates a list output containing the new dataframe, an additional new dataframe consisting of predicted curve fit values, the regression model, and the saturation curve plot.

```{r, fig.height = 4, fig.width = 5}
g_sat_curve(d_std_g, d_sat_g)
```
\

#### 3.3.3 Plot standard curve , Method 2
*s_std_curve()* plots spectral data vs standard concentration (only in presence of homogenate).  It reports linear model summary statistics. It creates a list output containing the summary statistics and standard curve plots.

```{r, fig.height = 4, fig.width = 5}
s_std_curve(d_std)
```
\

#### 3.3.4 Plot saturation curve, Method 2
*s_sat_curve()* converts the spectral data to concentration of standard. The new dataframe contains the average slope (rate of reaction) and standard deviation for each replicate at each substrate concentration. *s_sat_curve* plots the new dataframe with substrate concentration on the x-axis, and rate of reaction on the y-axis. It predicts and reports Vmax and Km values. It creates a list output containing the new dataframe, an additional new dataframe consisting of predicted curve fit values, the regression model, and the saturation curve plot.

```{r, fig.height = 4, fig.width = 5}
s_sat_curve(d_std, d_sat)
```
\

#### 3.3.5 Plot raw saturation data, both methods
*sat_raw()* plots raw saturation curve data into separate facets based on substrate concentration. It creates a list output containing the raw data plot.

```{r, fig.height = 4, fig.width = 5}
sat_raw(d_sat)
```
\

#### 3.3.6 Add additional arguments to label axes and manually input starting values for Km and Vmax
Add the arguments, 'x.label =' and 'y.label =' to label the axes. Add the arguments, 'km =' and 'vmax =' to input estimation values for km and vmax. If these arguments are not defined, the functions will leave the labels blank and use its own system for estimating km and vmax.

```{r, eval = FALSE}
g_std_curve(d_std_g, x.label = "x-axis label", y.label = "y-axis label")
```

```{r, eval = FALSE}
s_std_curve(d_std_g, x.label = "x-axis label", y.label = "y-axis label")
```

```{r, eval = FALSE}
g_sat_curve(d_std, d_sat, x.label = "x-axis label", y.label = "y-axis label", km = 97, vmax = 0.03)
```

```{r, eval = FALSE}
s_sat_curve(d_std, d_sat, x.label = "x-axis label", y.label = "y-axis label", km = 97, vmax = 0.03)
```

```{r, eval = FALSE}
sat_raw(d_sat, x.label = "x-axis label", y.label = "y-axis label")
sat_raw(d_sat_g, x.label = "x-axis label", y.label = "y-axis label")
```

## 4 References
German, D.P., Weintraub, M.N., Grandy, A.S., Lauber, C.L., Rinkes, Z.L., and Allison, S.D. Optimization of hydrolytic and oxidative enzyme methods for ecosystem studies. Soil Biology and Biochemistry 43(7), 1387-1397 (2011) doi:10.1016/j.soilbio.2011.03.017

Sinsabaugh, R.L., Belnap, J., Findlay, S.G. et al. Extracellular enzyme kinetics scale with resource availability. Biogeochemistry 121, 287–304 (2014) doi:10.1007/s10533-014-0030-y
    
Steen, A.D. and Arnosti, C. Long lifetimes of β-glucosidase, leucine aminopeptidase, and phosphatase in Arctic seawater. Marine Chemistry 123(1-4), 127-132 (2015) doi:10.1016/j.marchem.2010.10.006


